---
title: "Simple Example"
author: "Julian Stander"
date: "04/02/2018"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#", message = FALSE, warning = FALSE) 
```

```{r set-options, echo = FALSE, cache = FALSE}
options(width = 60)
```


# Comparing the Underlying Means of Two Groups

## Comparing the Underlying Means of Two Groups

```{r, echo = FALSE}
setwd("/Users/julianstander/Dropbox/MATH500/Practicals/Practical_9_t_test_ANOVA")
#
library(readr)
#
qd <- read_csv("MATH500_Questionnaire_Data.csv")
#
library(ggplot2)
#
ggplot(qd, aes(x = Sex, y = Height, fill = Sex)) + 
    geom_boxplot(varwidth = TRUE) +
    labs(x = "Gender", y = "Height (cms)") +
  theme(axis.title = element_text(size = 20, color = "black"), # Customize the titles of both axes 
        axis.text = element_text(size = 18, color = "black"), # Customize the text of both axes
        legend.title = element_text(size = 20, color = "black"),
        legend.text = element_text(size = 18, color = "black")) # Customize the facet text
```


## Comparing the Underlying Means of Two Groups, continued

We can ask: 

- Is there an \textcolor{red}{underlying difference in mean} height between females and males?

This is a __profound question__.  We are not asking whether there is a difference
between the mean height of females and the mean height of males in the
data collected. 

We are asking, _more generally_, whether there is
a difference between the mean height of females and the mean height of males in
a much bigger _population_.


## $t$-test: Underlying Difference in Mean

To answer this question, we can use a $t$-test:
```{r}
t.test(Height ~ Sex, data = qd, var.equal = TRUE)
```

## $p$-value: Underlying Difference in Mean

We need to focus on the $p$-value

```{r, echo = FALSE}
p_value <- t.test(Height ~ Sex, data = qd, var.equal = TRUE)$p.value
p_value
```

In general, if the $p$-value is less that $0.05$, 
\textit{there is an underlying difference in mean height between females and males}.

In this case, the $p$-value is considerably less than $0.05$. Therefore, we conclude that 

- \textbf{the data supply evidence that there is an underlying difference in mean height between females and males}.

## Obtaining the $p$-value using the \texttt{lm} Function

We can obtain the above $p$-value `r format(p_value, scientific = FALSE)` in a different way 

- using the \texttt{lm} function for \texttt{l}inear \texttt{m}odelling:

```{r, results = 'hide'}
m <- lm(Height ~ Sex, data = qd)
summary(m)
```

```{r, echo = FALSE}
summary(m)$coefficients
```

## Comparing the Underlying Means of Two Groups

We can also ask: 

- Is the mean height of females \textcolor{red}{less than} the mean height of males \textcolor{red}{in a wider population} of students?

Again, we are not asking whether the mean height of females is less than the mean height of males in the
data collected.  

We are asking, _more generally_, whether the mean height of females is less than the mean height of males in
a much bigger _population_.



## $t$-test: Is the Mean Height of Females Less Than the Mean Height of Males in the Population?

To answer this question, we can use a $t$-test, with 

\texttt{alternative = ``less''}:

\footnotesize
```{r}
t.test(Height ~ Sex, data = qd, var.equal = TRUE, 
       alternative = "less")
```

## $p$-value: Is the Mean Height of Females Less Than the Mean Height of Males in the Population?

We need to focus on the $p$-value

```{r, echo = FALSE}
t.test(Height ~ Sex, data = qd, var.equal = TRUE,
       alternative = "less")$p.value
```

If the $p$-value is less that $0.05$, we conclude that 

- \textbf{the mean height of females is less than the mean height of males in 
a wider population of students}.

- What do you conclude?

# Comparing the Underlying Means of More Than Two Groups

## Comparing the Underlying Means of More Than Two Groups: Some Data

Consider these data on the number of hours worked in a week by randomly selected accountants, GPs, lecturers and plumbers:

\begin{tabular}{l|ccccccccccc}\hline\hline
Accountants & 45& 38& 40& 42& 48& 37& 44& 40& 39& 42& 41\\
  & 40 & 36& 40& 48 & & & & & & & \\ \hline
GPs & 60& 57& 44& 52 & 57& 45& 42& 56& 53& 42 & 44\\
 &  54& 51& 58 & & & & & & & &\\ \hline
Lecturers & 52& 45& 40& 48& 36& 50& 56& 42& 37& 43& 47\\ \hline
Plumbers & 44& 39& 50& 37& 45& 39& 52& 45& 39& 48& 44\\
 & 43& 53 & & & & & & & & &\\ \hline \hline
\end{tabular}

## Comparing the Underlying Means of More Than Two Groups: Some Data, continued

A good way of storing these data in \R\ is to use a data frame with  one variable for the hours worked 
and another for the corresponding
employment group (accountants, GPs, lecturers or plumbers)

```{r, echo = FALSE}
accountants <- c(45, 38, 40, 42, 48, 37, 44, 40, 39, 42, 41, 40, 36, 40, 48) 
gps <- c(60, 57, 44, 52, 57, 45, 42, 56, 53, 42, 44, 54, 51, 58)
lecturers <- c(52, 45, 40, 48, 36, 50, 56, 42, 37, 43, 47) 
plumbers <- c(44, 39, 50, 37, 45, 39, 52, 45, 39, 48, 44, 43, 53)
n_a <- length(accountants)
n_g <- length(gps)
n_l <- length(lecturers)
n_p <- length(plumbers)
hours_worked <- c(accountants, gps, lecturers, plumbers)
group <- factor(c(rep("Accountants", n_a), 
                  rep("GPS", n_g), 
                  rep("Lecturers", n_l), 
                  rep("Plumbers", n_p)))
hw_g <- data.frame(hours_worked, group)
```

Here are some selected rows:

```{r}
hw_g[c(1, 4, 17, 22, 33, 39, 42, 48), ]
```

## Comparing the Underlying Means of More Than Two Groups: Plotting the Data

```{r, echo = FALSE}
ggplot(hw_g, aes(x = group, y = hours_worked, fill = group)) + 
    geom_boxplot(varwidth = TRUE) +
    labs(x = "Group", y = "Hours worked each week") +
  theme(axis.title = element_text(size = 20, color = "black"), # Customize the titles of both axes 
        axis.text = element_text(size = 18, color = "black"), # Customize the text of both axes
        legend.title = element_text(size = 20, color = "black"),
        legend.text = element_text(size = 18, color = "black")) # Customize the facet text
```


## One-way Analysis of Variance

- The plot displays the weekly hours worked by accountants, GPs, lecturers and 
plumbers.

- The question of interest is:
Is the underlying mean number of hours different between these \textbf{four} groups?

- To answer this question we perform a \textbf{one-way analysis of variance}. 

- \textbf{An}alysis \textbf{o}f \textbf{Va}riance is often abbreviated \textbf{ANOVA}.

## One-way Analysis of Variance using the \texttt{lm} Function
 
```{r}
m <- lm(hours_worked ~ group, data = hw_g) 
  # ANOVA is a linear model
anova(m)
```


## $p$-value 

Here, the $p$-value is

```{r, echo = FALSE}
anova(m)$"Pr(>F)"[1]
```
If the $p$-value is less that $0.05$, which it is here, we conclude that

- \textbf{there is a  difference between the 
underlying mean number of hours worked in these four groups}.


## One-way Analysis of Variance using the \texttt{aov} Function

A __One-way Analysis of Variance__ can also be performed using the \texttt{aov} function

```{r}
m_2 <- aov(hours_worked ~ group, data = hw_g)
summary(m_2)
```


## Follow-up Analysis

If we find that there is an  difference between the underlying means of the
groups, we should proceed by performing a __follow-up analysis__ to see where the group differences are:

\footnotesize
```{r}
TukeyHSD(m_2) # Follow-up: pair-wise comparisons
```


## Follow-up Analysis, continued

```{r, echo=FALSE}
options(scipen = 999)
```

```{r}
TukeyHSD(m_2)$group[,4]
```

- We should look at the $p$-values in the \texttt{p adj} column.

- When the $p$-value is
less that $0.05$, we should conclude that there is an underlying difference.  

- So here there are differences between GPs and accountants, between lecturers and GPs and between plumbers and GPs.  GPs work a lot!!!


